Optimizing the power of adaptive seamless designs using Bayesian Optimization

### 2020-11-23 main.tex
Updates from Overleaf

\section{Motivating Example}
\label{sect.motivation}

@@

% TODO: Jakob copy contents from paper
% Is there a condensed version? In Friede et al (2020) it is only explained in detail?
% TODO: Tim
Intro 3- 4 Sätze

\subsection{Two-stage design with treatment selection}
\label{ssec:two_stage_design_with_treatment_selection}

Multi-arm trial; two-stages; treatment selection in interim analysis; final analysis based on combination test approach with closed test principle

% first stage = interrim analysis
% second stage = final analysys
% stick to first second stage througout text but mention interrim here?

@@

\paragraph{Epsilion Rule}

@@

% TODO : Jakob

The goal of this simulation study is to investigate the potential benefit of the MBO approach for effectively finding a clinical trial design that maximizes a metric.
The analysis is motivated by the clinical example with COPD patients introduced in Section~\ref{sect.motivation}, with treatment selection in the first of two stages.
In the first stage five arms (defined by the control and four different administered doses) are considered, and in the second stage only a selected subset of arms is kept for further evaluation. 
The metric to be optimized is the statistical power, where a positive outcome means a trial arm with true effect is detected as significant effect.

@@

The strategies regarding treatment selection for the second stage in the two-phase trial design, following the concept of adaptive seamless designs discussed in Section~\ref{sec:adaptive_seamless_designs}, and their corresponding parameters that are used in the simulation study are given in Table~\ref{tab:selection_strategies} and Table~\ref{tab:search_space}, respectively.
The strategies 1, 2, 3, and \emph{all} select the $k$ best strategies according to their respective test statistics in the first stage (with \emph{all} representing all $k=4$ treatments besides the control). The strategies \emph{eps} and \emph{thresh} represent strategies with a flexible number of treatments taken to the second stage. 
In Table~\ref{tab:search_space} the ranges for these two parameters and for the proportion of patients (ratio $r$) allocated to the second stage are given.
\begin{table}[h]
  \caption{Overview of selection strategies to determine which trial arms are kept for second stage}
  \label{tab:selection_strategies}
  \centering
  \begin{tabular}{ll}
  \hline
  Notation       & Procedure of selection strategy  \\
  \hline
  \emph{all}     & Select all arms  \\
  $\{1,2,3\}$    & Select 1, 2 or 3 arms, respectively, according to maximal test statistic \\
  \emph{eps}     & Select all arms with test statistic less than $\epsilon$ (epsilon) below maximal test statistic \\
  \emph{thresh}  & Select all arms with test statistic larger than the value $\tau$ (threshold) \\
  \hline
  \end{tabular}
\end{table}
\begin{table}[h]
  \caption{Parameters for the selection strategies representing the search space of trial designs}
  \label{tab:search_space}
  \centering
  \begin{tabular}{ll}
  \hline
  Parameter                        & Range \\
  \hline
  Selection strategy               & $\{\text{all}, 1, 2, 3, \text{eps}, \text{thresh} \}$ \\
  $r$ (ratio)                      & $(0,1)$ \\
  $\epsilon$ (only for \emph{eps}) & $[0,4]$ \\
  $\tau$ (onlt for \emph{thresh})  & $[0,10]$ \\
  \hline
  \end{tabular}
\end{table}

%FIXME: JR: uniform naming: eps, thersh for the values and epsilon and treshold for the strategies

@@

Let $n_{\text{treat}}$ be the total number of treatments in the overall trial, and let $n_{\text{stage1}}$ and $n_{\text{stage2}}$ be the numbers of patients allocated to each treatment in stage 1 and stage 2 of the trial, respectively. Let $k_1$ and $k_2$ denote the number of treatments in stage 1 and stage 2, respectively. Within one stage, the same number of patients are allocated to all different treatments. Then $n_{\text{treat}}$ is given by
\begin{equation}
  \label{eq:ntreat}
  n_{\text{treat}} = k_1 \cdot n_{\text{stage1}} + k_2 \cdot n_{\text{stage2}}.
\end{equation}
In our scenario (as described in Section \ref{subsec:design}) we have $k_1 = 5$ and $k_2 \in \{2, \ldots, 5\}$, where $k_2$ depends on the result of the treatment selection strategy in stage 1 and the control treatment is included in both stages.

@@

The calibration step is necessary to guarantee that the total number of treatments $n_{\text{treat}}$ is (in average) the same, also for the variable selection strategies $\text{threshold}$ and $\text{epsilon}$. Then the black-box from formula (\ref{eq:bbox}) can be reformulated as 
\begin{equation}
  y = \tilde{f}(n_{\text{treat}}, r, \text{effect set}, \text{selection strategy}, \epsilon, \tau).
\end{equation}
In our simulations, $n_{\text{treat}}$ is considered to be a constant.

@@

In the simulation study we compare an exhaustive grid search on the parameter space with MBO (Model-based optimization) with respect to both performance (power) and runtime.

First, we describe the \emph{exhaustive grid search} (\emph{Grid}).
For each effect set (see Table~\ref{tab:table_effect_names}) and for each selection strategy, we conduct an exhaustive grid search with a resolution of 25 points per real-valued dimension, i.e.\ one parameter ($r$) for the strategies all, 1, 2, and 3, and two parameters ($r$ and $\epsilon$ or $\tau$, respectively).
This results in a total of $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for each effect set.
Additionally, we will use a subset of this grid of 7 points per dimensions, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$ design configurations, which we will call \emph{Grid Small}.
Both grids are evaluated with 20 stochastic replications to estimate the variance of the estimated power.

The best point per evaluated grid is determined by selecting the configuration (parameter set) with the best outcome, i.e.\ the largest power.
The corresponding value is generally too optimistic, as the random noise on the estimated value can contribute to the good outcome.
To account for this optimism and report unbiased performance estimates, for each of the 20 replicate simulations, we identify the best configuration and consider the corresponding 19 outcomes of the other 19 replicate simulations, similar to a cross-validation framework in machine learning applications. 
%This eliminates the risk of only reporting outcomes that are optimal by chance.

Next, we describe our realization of the search with the \emph{MBO (Model-based optimization)} approach.
For the application of MBO, we first have to define how the surrogate model that predicts the outcome of $\tilde{f}$ for unknown values of $\theta$ (parameter sets) is generated.
As the regression model for the surrogate we choose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is known to work well for fairly low dimensional search spaces.
We configure Kriging to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.

Kriging expects a numerical input, but our search space contains categorical parameters and is even hierarchical.
First, to transform categorical parameters of the search space (see Table~\ref{tab:search_space}) into numerical values we use one-hot encoding.
Furthermore, our search space is hierarchical because the parameters $\epsilon$ and $\tau$ are only active for the respective selection strategies \emph{eps} and \emph{thresh}.
The hierarchical structure introduces inactive values for the configurations ($\theta$), i.e.\ if the \emph{selection strategy} is set to \emph{eps} the parameter $\tau$ of \emph{thresh} is inactive.
If a value is inactive, we set it to a value two times as high as the maximum in its active range (e.g.\ if $\tau$ is inactive it will be set to 20).
This ensures that within the original range the Kriging model is not or only minimally affected by the inactive values.
Note, that this trick only works because we did not scale the input variables to $[0,1]$ and because the estimated covariance of the Mattern$\frac{5}{2}$ kernel is approximately zero for distances greater than $4$.
With the above steps we ensure that the input for the Kriging is purely numeric.

@@

At the end of the optimization procedure, the best configuration is chosen according to the best prediction obtained on the set of evaluated points as explained in Section~\ref{ssec:best_point}.
Again, overoptimism must be avoided. 
Thus, the best selected configuration (the best parameter set) is then independently evaluated $10$ times again.

The whole MBO procedure is repeated $10$ times to ensure that a single optimization run was not just lucky in finding the optimum.

Additionally we want to investigate whether the results of \emph{MBO} are better then \emph{grid search} because \emph{MBO} can propose configurations within the whole search space and not only limited the grid.
Therefore we introduce \textbf{MBO Grid}.
The optimization is executed as described for \emph{MBO} but with one difference:
At the end we conduct the independent evaluation for the configuration within the \emph{grid} that is closest to the optimal configuration that is proposed by \emph{MBO}.
% Mention that only configurations of the same factor level are looked at?

We expect that \emph{MBO} achieves to find configurations that yield better results than the \emph{grid search} for two reasons.
First, \emph{MBO} will be able to find an optimal configuration that is not restricted to be within the grid but can be between points from the grid.
Second, \emph{MBO} will not choose configurations that were observed to perform well only due to noise.

@@

In this chapter we present the results of the unbiased comparison on independent replicates for the exhaustive grid search, the small grid search and the MBO procedure.
Figure~\ref{fig:plot_boxplot_valid_y} shows the  estimated power values $y$ for the best solutions of the three approaches, for the four different effect sets and for three different numbers of $n_{\text{treat}}$.
\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y.pdf}
\caption{%
  Boxplots of performance (power values) on independent replicates of the best solutions for the four approaches, the different effect sets, and different numbers of $n_{\text{treat}}$. %FIXME rename n_cases in plot
  }
\label{fig:plot_boxplot_valid_y}
\end{figure}

The plots show that \emph{MBO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method, for all effects sets. Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$).
In other cases \emph{MBO} yields slightly better results (\emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
Across all scenarios, \emph{MBO} and \emph{Grid} perform similar, even though \emph{MBO} due to its evaluation budget is allowed only 116 evaluations in contrast to the 1350 evaluations for \emph{Grid}, with a resolution of 25 points per dimension. 
The difference in the number of evaluations is clearly reflected in the average runtime of the optimization methods given in Table~\ref{tab:table_time}. \emph{MBO} is approximately 20 times faster.
\input{generated/tables/table_time.tex}

For \emph{MBO Grid}, only solutions on the grid that was evaluated by \emph{Grid} are considered in the evaluation. 
Comparing the results of \emph{MBO Grid} to those of \emph{MBO} and the exhaustive \emph{Grid}, we can observe two patterns.
First, if \emph{Grid} achieves equal or better results then \emph{MBO}, the results of \emph{MBO Grid} are similar to \emph{MBO}. 
If \emph{MBO} was not able to find a better solution then \emph{Grid}, then also \emph{MBO Grid} cannot yield better results as it just takes the configuration on the Grid that is closest to the configuration found by \emph{MBO}.
An increase in performance of \emph{MBO Grid} compared to \emph{MBO} is purely due to chance.
Second, and in contrast to the first pattern: 
If \emph{MBO} is able to find better configurations then the exhaustive \emph{Grid} (e.g. for scenario \ \emph{effect: linear}, $n_{\text{treat}} = 1000$), then also the results of \emph{MBO Grid} are better then the results of \emph{Grid}.

This leads to the conclusion that the resolution of the exhaustive grid is sufficiently fine but the selection of the final configuration within the grid search is not good enough in such scenarios.
The grid search has no notion of the stochasticity of the problem, i.e. in this case the noise on the predicted power. 
Just the configuration on the grid with the best outcome is "optimistically" selected as final configuration.
The independent validation then is based on $y_{\text{valid}}$ that represents the unbiased performance of this final configuration, and the overoptimism becomes apparent.
\emph{MBO} performs better, since the final configuration is determined based on the mean prediction of the surrogate of \emph{MBO}, as explained in Section~\ref{ssec:best_point}.
%FIXME: Plot y vs y_valid?
Below we will study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of \emph{Grid}. %FIXME: n_treat 500, paper 2 would be more interesting or n_treat 1000 and linear)

To compare \emph{MBO} with a grid search that uses approximately the same time (see~\ref{tab:table_time}) and number of evaluations, we consider \emph{Grid Small}.
Here, a resolution of 7 points per dimension leads to 126 evaluations, close to 116 allowed evaluations of \emph{MBO} (see~\ref{tab:table_time}), with only a small advantage for \emph{Grid Small}.
Despite the 10 more allowed evaluations, in the majority of cases the results of \emph{Grid Small} are inferior to the results of \emph{MBO} as can be seen in Figure~\ref{fig:plot_boxplot_valid_y}.
Only for the scenarios (\emph{effect: paper}, $n_{\text{treat}} = 2000$) and (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$) \emph{Grid Small} has a small advantage over \emph{MBO} and \emph{Grid}.
In the first case, the absolute differences are negligible as all methods lead to a high power.
For the second scenario we will later show, that the optimal configuration is close to the border of the search space.
The higher resolution of the exhaustive \emph{Grid} probably leads to the selection of slightly suboptimal configurations close to the optimum due to the stochasticity.
These points are not included in the coarse grid.
In other words, if we are lucky and \emph{Grid Small} covers the optimum we are more likely to select it correctly because of less competing candidates and in consequence less overoptimism.

@@

%After having looked at the best obtained outcome on our problem scenarios and established the ability of \emph{MBO} to find equally good configurations as the much more demanding exhaustive \emph{Grid}
We now investigate the results in more detail.
In Figure~\ref{fig:plot_allbest} the observed outcomes during the evaluation of the grid as well as the final configurations found by \emph{MBO} are shown, in order to analyze the differences in dependence of the parameters \emph{selection strategy} and \emph{stage ratio} ($r$).
\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{generated/figures/plot_allbest.pdf}
\caption{%
  All evaluated configurations and their outcomes on the \emph{Grid}.
  For the epsilon and threshold rule only the curve for the $\epsilon$ and $\tau$ value that reached the best power ($y$) is displayed.
  For each configuration in the grid 10 stochastic results are obtained.
  A line connects the mean outcomes for each configuration.
  Additionally the best found configurations from \emph{MBO} are displayed.
  }
\label{fig:plot_allbest}
\end{figure}
Note that the \emph{stage ratio} is not drawn directly but the resulting number of treatments in the first stage.
For the \emph{selection strategies} \emph{epsilon rule} and \emph{threshold rule} only curves representing the $\epsilon$ and $\tau$ values with the best outcome, respectively, are shown.
Note that for the effect sets \emph{linear} and \emph{sigmoid} the curves for selection strategy \emph{1 best} and \emph{eps} are as expected nearly identical.
As intuitively anticipated, in almost all scenarios the worst results were obtained if all available treatments are used in stage 1 and none in stage 2. 
In all scenarios, \emph{MBO} identified the peak of all combined curves.
Within the 10 stochastic repetitions per scenario, \emph{MBO} found different configurations, but those had similarly good outcomes.
That no unique best setting was found across the repetitions can be attributed to the relatively high noise of the simulation and the relatively flat peaks, i.e. different configurations yield similarly good outcomes.
In cases in which multiple optimization runs result in different configurations it might be advisable to let a human expert choose the final configuration, also depending on the medical background.
%FIXME: Write more about the meaning of the results from a clinical point of view.

>>>>

\section{Motivating Example}
\label{sec:motivating_example}

@@

% TODO: Jakob copy contents from paper
% Is there a condensed version? In Friede et al (2020) it is only explained in detail?
% TODO: Tim
Traditionally drug development is organized in four phases with individual clinical trials for the separate phases. Seamless designs combine elements from different development phases in a single trial thereby offering the promise of speeding up development times. For instance, seamless phase II/III designs combine the learning about dose-response relationship or subgroup heterogeneity typical for phase II with confirmation of treatment effects in  phase III. If the data of the phase II part are used for confirmatory purposes, then this needs to be accounted for in the testing strategy to maintain the familywise type I error rate in the strong sense. This type of designs are referred to as adaptive seamless designs. A recent overview can be found in \citet{friede_adaptive_2020}.

\subsection{Two-stage design with treatment selection}
\label{ssec:two_stage_design_with_treatment_selection}
Here we consider a particular type of adaptive seamless designs, namely two-stage designs with treatment (or more specifically dose) selection. The trial starts in stage 1 with $K$ experimental treatments or doses of one experimental treatment which are compared to a common control. Interest is in assessing the treatment effects $\theta_k$ with $k=1,\dots, K$ comparing experimental treatment $k$ vs. control and testing the null hypotheses $H_{0,k}: \theta_k=0$ against the alternative  $H_{1,k}: \theta_k>0$ (larger $\theta_k$ being better). Fur this purpose test statistics $Z_{1,k}$ based on data from the first design stage are considered with larger $Z_{1,k}$ indicating more beneficial treatments. In an interim analysis, treatments that are continued into the second stage are selected; the set of selected treatments is denoted by $\mathcal{S}$. The selection rules used here are described in Section \ref{ssec:selection_rules}. Based on the data from patients recruited into the second stage, test statistics $Z_{2,k}$ with $k \in \mathcal{S}$ are computed.

final analysis based on combination test approach with closed test principle

% first stage = interim analysis
% second stage = final analysis
% stick to first second stage throughout text but mention interim here?

@@

\paragraph{Epsilon Rule}

@@

The goal of this simulation study is to investigate the potential benefit of MBO for effectively finding a clinical trial design that maximizes a metric.
The analysis is motivated by the clinical example with COPD patients introduced in Section~\ref{sec:motivating_example}, with treatment selection in the first of two stages.
In the first stage five arms (defined by the control and four different administered doses) are considered, and in the second stage only a selected subset of arms is kept for further evaluation. 
The metric to be optimized is the statistical power, where a positive outcome means a trial arm with true effect is detected as significant effect.

@@

The strategies regarding treatment selection for the second stage in the two-phase trial design, following the concept of adaptive seamless designs discussed in Section~\ref{sec:adaptive_seamless_designs}, and their corresponding parameters that are used in the simulation study are given in Table~\ref{tab:selection_strategies} and Table~\ref{tab:search_space}, respectively.
The strategies 1, 2, 3, and \emph{all} select the $k$ best strategies according to their respective test statistics in the first stage (with \emph{all} representing all $k=4$ treatments besides the control). 
The strategies \emph{eps} and \emph{thresh} represent strategies with a flexible number of treatments taken to the second stage. 
In Table~\ref{tab:search_space} the ranges for these two parameters and for the proportion of patients (ratio $r$) allocated to the second stage are given.
\begin{table}[h]
  \caption{Overview of selection strategies to determine which trial arms are kept for second stage}
  \label{tab:selection_strategies}
  \centering
  \begin{tabular}{ll}
  \hline
  Notation       & Procedure of selection strategy  \\
  \hline
  \emph{all}     & Select all arms  \\
  $\{1,2,3\}$    & Select 1, 2 or 3 arms, respectively, according to maximal test statistic \\
  \emph{eps}     & Select all arms with test statistic less than $\epsilon$ (epsilon) below maximal test statistic \\
  \emph{thresh}  & Select all arms with test statistic larger than the value $\tau$ (threshold) \\
  \hline
  \end{tabular}
\end{table}
\begin{table}[h]
  \caption{Parameters for the selection strategies representing the search space of trial designs}
  \label{tab:search_space}
  \centering
  \begin{tabular}{ll}
  \hline
  Parameter                        & Range \\
  \hline
  Selection strategy               & $\{\text{all}, 1, 2, 3, \text{eps}, \text{thresh} \}$ \\
  $r$ (ratio)                      & $(0,1)$ \\
  $\epsilon$ (only for \emph{eps}) & $[0,4]$ \\
  $\tau$ (only for \emph{thresh})  & $[0,10]$ \\
  \hline
  \end{tabular}
\end{table}

@@

Let $n_{\text{treat}}$ be the total number of treatments in the overall trial, and let $n_{\text{stage1}}$ and $n_{\text{stage2}}$ be the numbers of patients allocated to each treatment in stage 1 and stage 2 of the trial, respectively. 
Let $k_1$ and $k_2$ denote the number of treatments in stage 1 and stage 2, respectively. 
Within one stage, the same number of patients are allocated to all different treatments. 
Then $n_{\text{treat}}$ is given by
\begin{equation}
  \label{eq:ntreat}
  n_{\text{treat}} = k_1 \cdot n_{\text{stage1}} + k_2 \cdot n_{\text{stage2}}.
\end{equation}
In our scenario (as described in Section \ref{subsec:design}) we have $k_1 = 5$ and $k_2 \in \{2, \ldots, 5\}$, where $k_2$ depends on the result of the treatment selection strategy in stage 1.
The control treatment is included in both stages.

@@

The calibration step is necessary to guarantee that the total number of treatments $n_{\text{treat}}$ is (in average) the same, also for the variable selection strategies \emph{eps} and \emph{thresh}.
Then the black-box from formula (\ref{eq:bbox}) can be reformulated as 
\begin{equation}
  y = \tilde{f}(n_{\text{treat}}, r, \text{effect set}, \text{selection strategy}, \epsilon, \tau).
\end{equation}
In our simulations, $n_{\text{treat}}$ is considered to be a constant.

@@

In the simulation study we compare the ability to two optimization approaches for finding an optimal configuration, on the one hand an exhaustive grid search on the parameter space and on the other hand  MBO (Model-based optimization).
For different scenarios we compare the statistical power of the best found clinical trial design.
We also compare the runtime needed to determine the respective optimization results.
Next we introduce the two optimization strategies.

\paragraph{Grid}
We conduct an \emph{exhaustive grid search} with a resolution of 25 points per real-valued dimension and one point for each categorical dimension.
Referring to our application, for each selection strategy we evaluate a numerical grid with a resolution of 25.
All 6 selection strategies have the numeric parameter $r$.
The strategies \emph{epsilon} and \emph{thresh} both have an additional second parameter, $\epsilon$ or $\tau$, respectively. 
This results in a total of $4 \cdot 25 + 2 \cdot 25 \cdot 25 = 1350$ design configurations that are evaluated for one scenario.
We call this approach \textbf{Grid}.
Further, we also consider a subset of this grid of 7 points per dimensions, resulting in $4 \cdot 7 + 2 \cdot 7 \cdot 7 = 126$ design configurations, which we will call \textbf{Grid Small}.
Both grids are evaluated with 20 stochastic replications to estimate the variance of the estimated power.

The best point per evaluated grid is determined by selecting the configuration with the best outcome, i.e.\  the parameter set with the largest power.
The corresponding value is generally too optimistic, as the random noise on the estimated value can contribute to the quality of the outcome.
To account for this optimism and report unbiased performance estimates, for each of the 20 replicate simulations, we identify the best configuration and consider the corresponding 19 outcomes of the other 19 replicate simulations, similar to a cross-validation framework in machine learning applications. 
%This eliminates the risk of only reporting outcomes that are optimal by chance.

\paragraph{MBO (Model-based optimization)}
For applying MBO, we first define how the surrogate model that predicts the outcome of $\tilde{f}$ for unknown values of $\theta$ (parameter sets) is generated.
As regression model for the surrogate we choose Kriging (also called Gaussian process regression) and use the implementation in the R-package \texttt{DiceKriging}~\citep{roustant_dicekriging_2012}, because it is known to work well for fairly low dimensional search spaces.
We configure Kriging to apply the Mattern$\frac{5}{2}$ kernel with an estimated \emph{nugget effect} to account for the noisy response of $f$ and without scaling the input variables to $[0,1]$.

Kriging expects a numerical input, but our search space contains categorical parameters and is even hierarchical.
First, to transform the categorical parameters of the search space (see Table~\ref{tab:search_space}) into numerical values we use one-hot encoding.
Furthermore, our search space is hierarchical because the parameters $\epsilon$ and $\tau$ are only active for the respective selection strategies \emph{eps} and \emph{thresh}.
The hierarchical structure introduces inactive values for the configurations ($\theta$), i.e.\ if \emph{selection strategy} is set to \emph{eps} then the parameter $\tau$ of \emph{thresh} is inactive.
If a value is inactive, we set it to a value two times as high as the maximum in its active range (e.g.\ if $\tau$ is inactive it will be set to 20).
This ensures that within the original range the Kriging model is not or only minimally affected by the inactive values.
Note, that this trick only works because we did not scale the input variables to $[0,1]$ and because the estimated covariance of the Mattern$\frac{5}{2}$ kernel is approximately zero for distances greater than $4$.
With the above steps we ensure that the input for the Kriging is purely numeric.

@@

At the end of the optimization procedure, the best configuration is chosen according to the best prediction obtained on the set of evaluated points as explained in Section~\ref{ssec:best_point}.
Again, overoptimism must be avoided. 
Thus, the best selected configuration (the best parameter set) is then independently evaluated $20$ times again.

The whole MBO procedure is repeated $20$ times to ensure that a single optimization run was not just lucky in finding the optimum.

Additionally we investigate how \emph{MBO} performs when it is limited to return configurations that are within the grid.
Therefore we introduce \textbf{MBO Grid}, which executes the optimization exactly as \emph{MBO}, but the best configuration is mapped to the closest value on the grid. 
In other words, for \emph{MBO Grid} we conduct the independent evaluation using the configuration within the grid that is closest to the optimal configuration that is proposed by \emph{MBO}.
The comparison of the results of \emph{MBO Gird} and \emph{Gird} will allow two possible conclusions:
First, if \emph{MBO Grid} is worse then \emph{MBO}, we can assume that the resolution of \emph{Grid} is not high enough to find an optimal configuration.
Second, if \emph{MBO Grid} is better then \emph{Grid}, it shows that \emph{Grid} did not select the optimal configuration, due to noise.
Summing up, there are two factors that can prevent \emph{Grid} from finding the optimal configuration: lack of resolution and noise of the outcome.

@@

In this chapter we present the results of the four optimization strategies: exhaustive \emph{Grid} search, grid search with a reduced resolution (\emph{Grid Small}), \emph{MBO} and \emph{MBO Grid},
We focus our comparison on independent, replicated evaluations of the configurations proposed by each optimizer.
The replicated evaluations give us an unbiased estimate of the performance ($y_{\text{valid}}$) of each optimizer.

For this study we consider 12 scenarios.
The scenarios are generated by combining the four different effect sets (see Table~\ref{tab:table_effect_names}) with the three different numbers of $n_{\text{treat}}$.
Each optimizer is applied on each scenario.

Figure~\ref{fig:plot_boxplot_valid_y} shows the estimated power values $y_{\text{valid}}$ for the best solutions in each scenario.
\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{generated/figures/plot_boxplot_valid_y.pdf}
\caption{%
  Box plots of performance (power) values on independent replicates of the best solutions for the four optimizers, the different effect sets, and different numbers of $n_{\text{treat}}$. %FIXME rename n_cases in plot
  }
\label{fig:plot_boxplot_valid_y}
\end{figure}

The plots show that \emph{MBO} generally gives results that are comparable to those obtained with the (exhaustive) \emph{Grid} method, for all effects sets. 
Only in some cases \emph{Grid} has a slight advantage (\emph{effect: linear}, $n_{\text{treat}} = 2000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 2000$).
Below we will see that these scenarios have in common that the optimal configuration is close to the border of the search space. %FIXME Make sure that this is mentioned for figure 4!
In other cases \emph{MBO} yields slightly better results (\emph{effect: paper}, $n_{\text{treat}} = 2000$; \emph{effect: paper2}, $n_{\text{treat}} = 500$; \emph{effect: sigmoid}, $n_{\text{treat}} = 500$).
Across all scenarios, \emph{MBO} and \emph{Grid} perform similar, even though \emph{MBO} due to its evaluation budget is allowed only 116 evaluations in contrast to the 1350 evaluations for \emph{Grid}, with a resolution of 25 points per dimension. 
The difference in the number of evaluations is clearly reflected in the average runtime of the optimization methods given in Table~\ref{tab:table_time}.
\emph{MBO} is approximately 20 times faster.
\input{generated/tables/table_time.tex}

For \emph{MBO Grid}, only solutions on the grid that was evaluated by \emph{Grid} are considered in the evaluation. 
Comparing the results of \emph{MBO Grid} to those of \emph{MBO} we mainly see similar performances for both.
This indicates, that the resolution of \emph{Grid} is fine enough to achieve performance similar to \emph{MBO}.

Comparing \emph{MBO Grid} to \emph{Gird} we can observe, that for the scenarios \emph{effect: paper}, $n_{\text{treat}} = 500$ and \emph{effect: paper}, $n_{\text{treat}} = 1000$ \emph{MBO Grid} achieves slightly better performance then \emph{Grid}.
The \emph{Grid} optimization selected the best outcome "optimistically" from all observed outcome and has no notion of the stochasticity of the problem, i.e. in this case the noise on the predicted power. 
Accordingly, we assume that for the mentioned scenarios the noise was so high, that \emph{Gird} was misguided by overoptimistic outcomes when selecting the final configuration as optimization result.
The independent validation $y_{\text{valid}}$ represents the unbiased performance of this final configuration, and the overoptimism becomes apparent.
In these scenarios \emph{MBO} performs better, since the final configuration is determined based on the mean prediction of the surrogate of \emph{MBO}, as explained in Section~\ref{ssec:best_point}.
Below we will study on a single scenario if a reduced noise due to a higher value of \texttt{nsim} decreases the effect of the overly optimistic selection of \emph{Grid}. %FIXME: n_treat 500, paper 2 would be more interesting or n_treat 1000 and linear)

If \emph{MBO} was not able to find a better solution then \emph{Grid}, then also \emph{MBO Grid} cannot yield better results as it just takes the configuration on the Grid that is closest to the configuration found by \emph{MBO}.
An increase in performance of \emph{MBO Grid} compared to \emph{MBO} is purely due to chance.

To compare \emph{MBO} with a grid search that uses approximately the same time (see~\ref{tab:table_time}) and number of evaluations, we consider \emph{Grid Small}.
Here, a resolution of 7 points per dimension leads to 126 evaluations, close to 116 allowed evaluations of \emph{MBO} (see~\ref{tab:table_time}), with only a small advantage for \emph{Grid Small}.
Despite the 10 more allowed evaluations, in the majority of cases the results of \emph{Grid Small} are inferior to the results of \emph{MBO} as can be seen in Figure~\ref{fig:plot_boxplot_valid_y}.
Only for the scenarios (\emph{effect: paper2}, $n_{\text{treat}} = 1000$) and (\emph{effect: sigmoid}, $n_{\text{treat}} = 1000$) \emph{Grid Small} has a small advantage over \emph{MBO} and \emph{Grid}.
In the first case, the absolute differences are negligible as all methods lead to a high power.
For the second case we will later show, that the optimal configuration is close to the border of the search space.
The higher resolution of the exhaustive \emph{Grid} probably leads to the selection of slightly suboptimal configurations close to the optimum due to the stochasticity.
These points are not included in the coarse grid.
In other words, if we are lucky and \emph{Grid Small} covers the optimum we are more likely to select it correctly because of less competing candidates and in consequence less overoptimism.
The advantage of \emph{MBO} over \emph{Gird Small} becomes apparent for all scenarios with \emph{effect: linear} and scenario \emph{effect:paper2}, $n_{\text{treat}} = 500$.

@@

%After having looked at the best obtained outcome on our problem scenarios and established the ability of \emph{MBO} to find equally good configurations as the much more demanding exhaustive \emph{Grid}
We now investigate the results in more detail.
In Figure~\ref{fig:plot_allbest} the observed outcomes during the evaluation of the grid as well as the final configurations found by \emph{MBO} are shown, in order to analyze the differences in dependence of the parameters \emph{selection strategy} and \emph{stage ratio} ($r$).
\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{generated/figures/plot_allbest.pdf}
\caption{%
  All evaluated configurations and their outcomes on the \emph{Grid}.
  For the epsilon and threshold rule only the curve for the $\epsilon$ and $\tau$ value that reached the best power ($y$) is displayed.
  For each configuration in the grid 10 stochastic results are obtained.
  A line connects the mean outcomes for each configuration.
  Additionally the best found configurations from \emph{MBO} are displayed.
  }
\label{fig:plot_allbest}
\end{figure}
Note that the \emph{stage ratio} is not drawn directly but the resulting number of treatments in the first stage.
For the \emph{selection strategies} \emph{epsilon rule} and \emph{threshold rule} only curves representing the $\epsilon$ and $\tau$ values with the best outcome, respectively, are shown.
Note that for the effect sets \emph{linear} and \emph{sigmoid} the curves for selection strategy \emph{1 best} and \emph{eps} are as expected nearly identical.
As intuitively anticipated, in almost all scenarios the worst results were obtained if all available treatments are used in stage 1 and none in stage 2. 
In all scenarios, \emph{MBO} identified the peak of all combined curves.
Within the 20 stochastic repetitions per scenario, \emph{MBO} found different configurations, but those had similarly good outcomes.
That no unique best setting was found across the repetitions can be attributed to the relatively high noise of the simulation and the relatively flat peaks, i.e. different configurations yield similarly good outcomes.
In cases in which multiple optimization runs result in different configurations it might be advisable to let a human expert choose the final configuration, also depending on the medical background.
The curves also show the potential of optimization for the different scenarios.
If the allowed treatment number is high ($n_{\text{treat}} = 1000$), then nearly all design configurations yield high statistical power and optimization is less necessary.
If the allowed treatment number is too small $n_{\text{treat}} = 500$, no configuration yields a practically usable design.
Here, a fast optimization helps to determine if a desired treatment number can actually yield sufficiently powerful designs.
Of special interest are the scenarios where the correct choice of the design configuration can make a difference between a design with a usable power or a design that is too weak, e.g. for the scenarios (\emph{effect: linear}, $n_{\text{treat}} = 1000$; \emph{effect: linear}, $n_{\text{treat}} = 2000$, \emph{effect: paper}, $n_{\text{treat}} = 1000$; \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$).
Note, that the selection strategies \emph{thresh} and \emph{all} never gave powerful designs.

%FIXME: Write more about the meaning of the results from a clinical point of view.

%In contrast to Figure~\ref{fig:plot_allbest},
Figure~\ref{fig:plot_best_x} visualizes in detail which configurations were selected by \emph{MBO} and \emph{Grid} and the corresponding values for $y_{\text{valid}}$ (result of the independent validation).
\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{generated/figures/plot_best_x.pdf}
\caption{Best found $x$-values for \emph{Grid} and \emph{MBO} runs, i.e. corresponding values for $y_{\text{valid}}$ and ratio $r$ for the selected configurations}
\label{fig:plot_best_x}
\end{figure}
The optimal values for $\epsilon$ and $\theta$ are not shown.
For some scenarios most replications of \emph{Grid} returned the same optimal configurations (e.g.\ for \emph{effect: sigmoid}, $n_{\text{treat}} = 1000$).
In such cases the signal to noise ratio was high enough to reliably find the best configuration.
These are also those scenarios where \emph{Grid} was superior the \emph{MBO} as can be seen in the previously shown Box plots in Figure~\ref{fig:plot_boxplot_valid_y}.
Furthermore, for all scenarios with \emph{effect: sigmoid} and the scenarios \emph{effect: linear}, $n_{\text{treat}} = 500)$ and \emph{effect: paper}, $n_{\text{treat}} = 500)$, small values of $r$ lead to an optimal outcome.
Here, \emph{MBO} was not able to find these values close to the border of the search space.
A common remedy is to log-transform this value.
